---
title: "Housing Stock Choropleth"
output: github_document
---

First we'll load the packages.

```{r load packages, message=FALSE}
library(tidyverse)
library(sf)
library(showtext)
library(rayshader)
library(rayrender)
library(here)
# font_add("Space Grotesk", "~/Library/Fonts/SpaceGrotesk-Regular.ttf")
```

Then we'll load in our census data and turn it into a data table.

```{r}
tracts_df <- read_sf("data/SHP_chicago-tracts_housing-stock/acs2021_5yr_B25024_14000US17031010502.shp", quiet = TRUE)
```

But these column names aren't very descriptive and some are unnecessary.
We'll only select a few and rename them to be more helpful.

```{r}
tracts_cleaned <- tracts_df %>% 
  select(geoid, 
         name, 
         geometry, 
         total_units = B25024001, 
         one_detached = B25024002, 
         one_attached = B25024003, 
         two = B25024004, 
         three_or_four = B25024005, 
         five_to_nine = B25024006, 
         ten_to_nineteen = B25024007, 
         twenty_to_fortynine = B25024008, 
         fifty_plus = B25024009)
```

Then we'll calculate the average housing units per each tract.\n
(The weighting isn't going to be perfect. I only took the average value from each variable [e.g. (5 + 9)/2 = 7] but it should be good enough to see trends.)

```{r}
AVG_2 <- 2
AVG_3_4 <- 3.5
AVG_5_9 <- 7
AVG_10_19 <- 14.5
AVG_20_49 <- 34.5

tracts_average_units <- tracts_cleaned %>% 
  mutate(average_units = (one_detached + 
                          one_attached + 
                          (two * AVG_2) + 
                          (three_or_four * AVG_3_4) + 
                          (five_to_nine * AVG_5_9) + 
                          (ten_to_nineteen * AVG_10_19) + 
                          (twenty_to_fortynine * AVG_20_49) + 
                          (fifty_plus * 100) )/total_units,
         .after = total_units)
```

Finally we just have to plot it.

```{r fig.width = 10, fig.height = 6}
housing_units <- ggplot(tracts_average_units) +
  geom_sf(aes(fill = average_units),
              color = "black"
  ) +
  labs(
    title = "Average Housing Units per Residential Building",
    subtitle = "In Chicago census tracts.",
    fill = "Units",
    caption = "U.S. Census Bureau (2017-2021). Units in Structure American Community Survey 5-year estimates."
  ) +
  scale_fill_viridis_b(breaks = c(0, 3, 5, 10, 25, 50, 100), na.value = "gray10") +
  theme_void() +
  theme(
    plot.title = element_text(family = "Space Grotesk",
                              face = "bold",
                              size = 22.5, 
                              hjust = 0.5, 
                              color = "white"),
    plot.subtitle = element_text(family = "Space Grotesk",
                                 face = "bold",
                                 size = 16,
                                 hjust = 0.5,
                                 color = "gray50"),
    plot.background = element_rect(fill = "gray10", 
                                   color = NA),
    plot.caption = element_text(family = "Josefin Sans SemiBold",
                                color = "white",
                                size = 10,
                                hjust = 0.5),
    legend.text = element_text(family = "Josefin Sans SemiBold",
                               color = "white"),
    legend.title = element_text(family = "Josefin Sans SemiBold",
                                color = "white"),
    plot.margin = margin(10, 0, 5, 0)
  )
grid::grid.newpage()
grid::grid.draw(grid::rectGrob(gp = grid::gpar(fill = "gray10")))
print(housing_units, newpage = FALSE)
```

```{r}
bounding_box <- st_bbox(tracts_average_units) # order: xmin, ymin, xmax, ymax

# wasn't able to match ground_material with the plot.background
# so I made the limits huge so that you only see plot.background
rayshader_housing_units <- ggplot(tracts_average_units) +
  geom_sf(aes(fill = average_units),
           color = NA) +
  scale_fill_viridis_c() +
  coord_sf(
    xlim = c(bounding_box[1] - 0.25, bounding_box[3] + 0.25),
    ylim = c(bounding_box[2] - 0.25, bounding_box[4] + 0.25)
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.background = element_rect(fill = "gray15",
                                   color = "gray15")
  )

rgl::clear3d()

plot_gg(
  rayshader_housing_units,
  solid = FALSE,
  shadow = FALSE,
  scale = 100,
  theta = 330,
  phi = 45,
  zoom = 0.36
 )


render_highquality(
  filename = here("housing-stock_choropleth", "plots", "raytraced_10.png"),
  interactive = FALSE,
  preview = FALSE,
  # output settings
  width = 3000,
  height = 3000,
  samples = 500,
  min_variance = 5e-10,
  # light settings
  lightintensity = 1000,
  lightdirection = 290,
  lightaltitude = 38
)
```